Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-10-22 writeups/research_proposal/propsal.tex
presentations

---
title: "Comparison of MCMC algorithms in Stochastic Volatility Models"
author: 
    - "By Benjamin Wee (32921020)" 
    - "Supervised by Catherine Forbes and Lauren Kennedy"
# abstract: "The objective of this research is to validate and compare the computational methods used to estimate Bayesian stochastic volatility models. Specifically, Simulation Based Calibration will be used to identify biases and errors arising from the results of Markov Chain Monte Carlo (MCMC) algorithms. The source of any miscalibration include, but are not limited to, estimation strategies, choice of sampler, model specifications and coding mistakes. This study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches."
format:
  pdf:
    fig-pos: H
    include-in-header:
        text: |
          \usepackage{lipsum}
          \usepackage{setspace}
          \onehalfspacing
          \linespread{1.5}
fontsize: 12pt
geometry: margin=1in
mathspec: true
---

<!-- # Background and Motivation
Maybe focus on the MCMC as opposed to SV upfront.

## SV
Why model volatility. Quantifying risks, tail probabilities, worst case scenarios. These are all inputs in the pricing of assets and derivatives. (pg92 FE slides)

SV models is one way to model volatility. Unlike the classic Black Scholes option pricing formula, it assumes that the underlying variance of an asset's return is not constant. That is, the variance of the underlying asset return is treated as a latent random variable. This deals with some of the empirical short comings of the BS model such as the volatility smile and skew in the realised variance in financial assets.

SV models are typically expressed as non linear Gaussian state space models. This can make it difficult to estimate using classical methods. The likelihood is unavailable in closed form and there are at least as many variables as data points. There are however, a variety of Frequentist and Bayesian strategies for estimating SV models. Frequentist approaches include quasi likelihood estimation and GMM which rely on asymptotic sampling distributions. Bayesian approaches rely on computational Markov Chain Monte Carlo techniques to sample from the target joint posterior of these high dimensional state space models.

This research will focus on Bayesian estimation strategies for Stochastic Volatility models. Particular attention will be given to the computational tools and approaches used to sample from high dimensional posterior distributions. Early approaches for estimating these models relied on classic Gibbs and Metropolis Hastings algorithms to apply the MCMC routines. Since then, there have been many developments in the availability of new kinds of MCMC samplers.

Hamiltonian Monte Carlo is a . It has been popularised through its implementation in a variety of open source programming languages. It presents a new way to sample from increasingly complicated generative models .

However, as the developments of these tools enable the estimation of more complicated models (with relative ease and accessibility through the development of open source software), so do the risk of mistakes and the need for procedures diagnose problems or mistakes when implementing these methods. How do we know that these algorithms are doing what we expect them to be doing, and how can we effectively differentiate the strengths and weaknesses of different tools.

How do we validate the computation and implementation of our sampling strategies and algorithms in increasingly complex models? And how do these form part of the computational modelling workflow?    -->

<!-- ## HMC
- There have been many developemnts over the past few decades on estimating Bayesian models through Markov Chain Monte Carlo methods.

The original BUGs software implemented Gibbs and Metropolis Hastings MCMC algorithms for estimating this model.

In particular, the availability of Hamiltonian Monte Carlo, a variant of the Metropolis Hastings algorithm, represents a recent advancement MCMC.

- The development of HMC is another advancement in estimating complicated, high dimensional models. 
- Development of diagnostics for different types of HMC and tools for evaluating the convergence and efficiency of the MCMC sampler.

- SV models are a primary candidate for using MCMC methods? Difficult to

## SBC -->

<!-- and provide tools for evaluating other types models and algorithms.  -->

<!-- The objective of this research is to evaluate the performance of MCMC algorithms for estimating SV models through a simulation study. Simulations are useful since there is control over the data generating process. This will help us understand the behaviour of MCMC and evaluate its performance against known parameters. Particular attention will be given to the calibration of the MCMC algorithms in the stochastic volatility context. Further diagnostics to evaluate performance will be explored to determine the efficiency of the MCMC sampler.

The first step is to understand what "performance" of MCMC algorithms means and what exactly we are measuring. At a high level, the objective of MCMC is to sample from the target posterior distribution and to provide reliable inference.  Convergence to the target posterior can be measured through efficiency measures such as rhat and effective sample size.

- Explcitly outline what is compared. KSC and HMC algorithms and estimation strategies.  -->

<!-- 
Frequentist methods

- Variety of estimation strategies

- There are a variety of ways of estimating this model using a Bayesian frameworks.

Over the past few decades there have been many developments in computing hardware -->
<!-- 
RESEARCH GOAL: To compare the performance of different MCMC algorithms and estimation strategies when fitting stochastic volatility models. The study will use a discrete formulation is described  -->

<!-- The use of simulation and sampling are used as part of model diagnostics and evaluation. -->

## 1. Research Goal

@@

<!-- Stochastic volatility models are used to explicitly model the variance of returns in financial instruments and can be estimated using frequentist or Bayesian frameworks.  -->

<!-- 
The objective of this research is to validate the computational methods used to estimate Bayesian stochastic volatility models. This study will assess the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. A simulation study will compare the ability of these methods to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics to check these features are the effective sample size to check efficiency of the algorithm and tests of uniformity to determine calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy. -->

<!-- Simulation Based Calibration will be used to identify biases and errors arising from the results of Markov Chain Monte Carlo (MCMC) algorithms. The source of any miscalibration include, but are not limited to, estimation strategies, choice of sampler, model specifications and coding mistakes. This study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches. -->

<!-- The objective of this research is to validate and compare the computational methods used to estimate Bayesian stochastic volatility models. Specifically, Simulation Based Calibration will be used to identify biases and errors arising from the results of Markov Chain Monte Carlo (MCMC) algorithms. The source of any miscalibration include, but are not limited to, estimation strategies, choice of sampler, model specifications and coding mistakes. This study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches. -->

<!-- Specifically, they propose a method of sampling by using an offset mixture normal approximation with a reweighting procedure to sample from the posterior distribution. By linearing the model and using a mixture of Gaussians, they were able to apply a Kalman Filter to estimate the posterior density over the latent volatilities and static parameters. -->

<!-- The univariate stochastic volatility model described in their paper outlines the state space representation of the the model which is described below.
 -->
# 2. Background and Literature Review

## 2.1 Stochastic Volatility
Stochastic volatility models are used in financial econometrics to model the variance or volatility of returns in a financial instrument. The return of a financial asset often exhibits patterns in the variance which can be modeled to improve prediction and estimates of risk or tail events. Unlike other volatility models such as ARCH or GARCH, stochastic volatility models treat the variance as a random variable. Furthermore, these models have a state space representation since the variance is expressed as a latent stochastic process. State space models can be difficult to estimate since there is a latent state for every data observation in the time series, which means there are often more parameters than data points. These models can be estimated using frequent frameworks but this research will focus on Bayesian estimation strategies.

Kim, Shephard and Chib (1998) estimate a discrete time, univariate stochastic volatility model using Bayesian methods which is outlined below. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

$$
\begin{aligned}
y_t =& \space \beta exp(h_t/2) \epsilon_t \\
h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
h_1 \sim& \space normal\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
\end{aligned}
$$

$$
\begin{aligned}
\epsilon_t \sim& \space normal(0,1) \\
\delta_t \sim& \space normal(0,1)
\end{aligned}
$$

Setting $\beta=1$, the model can be expressed more succinctly as:

$$
\begin{aligned}
y_t \sim& \space normal(0, exp(h_t/2)) \\ 
h_1 \sim& \space normal \left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
h_{t+1} \sim& \space normal(\mu +\phi(h_t - \mu) , \sigma_{\eta}^2), \space\space t\neq 1\\ 
\end{aligned}
$$

Priors for the static parameters are defined below with conjugate priors on $\mu$ and $\sigma^2$:

$$
\begin{aligned}
\mu \sim& \space normal(0, 10^2) \\
\sigma_{\eta}^2 \sim& \space IG(5/2, (0.01\times 5) / 2) \\
\phi^{\ast} \sim& \space beta(20, 1.5) \\
\phi &=  2\phi^{\ast} - 1
\end{aligned}
$$

The prior on $\phi$ is a "stretched" beta distribution. This is a beta distribution (as defined on the parameter $\phi^*$) which has been transformed to have support (-1, 1).

### Estimation strategy
Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother[^1] (de Jong and Shephard, 1995).

[^1]: In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.

The standard Kalman Filter and smoother is used to compute the posterior distribution over the latent states. This requires the state and measurement equations to be linear and conditionally Gaussian. Since the relationship between $y_t$ and $h_t$ in the measurement equation is not linear, a transformation is applied by squaring and taking the log of $y_t$.

$$
\begin{aligned}
y_t^{*} &= log(y_t^2) \\ 
&= log((\epsilon_t exp(h_t/2))^2) \\
&=  log(exp(h_t)) + log(\epsilon_t^2) \\
&= h_t + log(\epsilon_t^2)  \\
&= h_t + z_t \\
\end{aligned}
$$

Where $z_t = log(\epsilon_t^2)$ follows a log chi-squared distribution with mean -1.2704 and variance 4.93. The relationship between $y_t$ and $h_t$ is now linear; however, the error is not Gaussian. Since it is not simple to sample from this parameterisation of the model, KSC use a mixture of Gaussians to **approximate** the first 4 moments of the log chi squared distribution. This is defined by:

$$
\begin{aligned}
f(z_t) = \sum_{i=1}^{K} q_if_N(z_i|m_i-1.2704, \nu_i^2)
\end{aligned}
$$

Where K is the mixture of normal densities $f_N$, component probabilities $q_i$, mean $m_i-1.2704$ and variance $\nu_i^2$. These parameters were selected using moment matching where they found 7 normal densities with varying mean and variance parameters best approximated the log chi squared moments. These parameters and weights can be found in Appendix A.

The model can be sampled via the Kalman Filter and simulation smoother since the model is now linear and conditionally gaussian. The static parameters $\mu$ and $\sigma^2$ are sampled directly from their conjugate posterior distributions whereas $\phi$ is sampled via a Metropolis Hastings accept/reject procedure. The details can be around in Appendix B.

## 2.2 Hamiltoninan Monte Carlo
Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature (Duane et al. 1987) before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks (Neal, 1995) and statistical computing (Neal 2011). The algorithm has since become widely available through open source development projects such as Stan (Stan Development Team, 2023) and PyMC (2023).

@@

The Stan programming language's implementation of Hamiltonian Monte Carlo will be used for this study. Stan's default algorithm, the No-U-Turn Sampler (Hoffman & Gelman 2014), allows for direct sampling of the specified stochastic volatility model. Hamiltonian Monte Carlo allows for sampling of the generative model and can flexibly handle complicated likelihood functions. This approach will also use the same priors as specified in the Gaussian mixture approximation.

## 2.3 Research contribution
The goal of this research is to determine which algorithm is better at estimating the discrete time stochastic volatility model (conditional on the same priors). My main contribution is to design a simulation study that compares these algorithms and determines which sampling strategy is better based on the calibration of posterior estimates and the efficiency of the MCMC sampler.

<!-- New developments in MCMC algorithms and improved computing resources have enabled estimation of richer and more complicated models. However, as the development of tools enables us to fit more complicated models, so does the need to validate the output of these tools. Cook, Gelman and Rubin (2006) introduce simulation techniques to evaluate the "correctness" of the software used to sample Bayesian models. Their contribution identified that cumulative density quantiles from the posterior samples can be tested for uniformity. Talts et al (2018). build upon this work by developing Simulation Based Calibration to provide more robust validation of output from a MCMC sampler through the use of rank statistics. -->

 <!-- through the use of the CDF quantiles approximated from the posterior distribution. -->

<!-- , through open source programming languages such as Stan and PyMC, have become

As new and complicated

There have been many developments in MCMC algorithms since the introduction of classical methods such as Gibbs and Metropolis Hastings.

Hamiltoninn Monte Carlo (algorithm introduced when, widely available since 2010...?) is a relatively new im.

- Since then, there have been many developments in MCMC algorithms. In particular, HMC provides a flexible alternative to fitting and sampling from complicated high dimensional Bayesian models.
- It's key innovation is the use of gradients - enabling the flexible specification of any likelihood function. 
- Constraints- can only sample from continuous 
- HMC also provides new diagnostics for evaluating the effectiveness of the sampling procedure and how well the sampler is targeting the joint dist.

- How do we compare models? SBC. ESS in the tails, rhats.

- Given this context, my research looks to compare the efficacy of KSCs proposed MCMC sampling strategy with Hamiltonina Monte Carlo (NUTs).  -->

# 3. Methodology

## 3.1 Design
Simulation Based Calibration (SBC) is a technique that checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples conditional on the simulated data.

To illustrate this procedure, let $\theta$ be a parameter and $y$ represent the dataset. Start with a single draw from the prior distribution:
$$
\begin{aligned}
\theta^{sim} \sim \pi(\theta)
\end{aligned}
$$

@@

$$
\begin{aligned}
y^{sim} \sim \pi (y|\theta^{sim})
\end{aligned}
$$

@@

$$
\begin{aligned}
\{\theta_1,\dots , \theta_{L}\} \sim \pi (\theta | y^{sim})
\end{aligned}
$$

A key result is that the posterior sample $\{\theta_1,\dots , \theta_{L}\}$ will share the same distribution as the prior samples $\theta^{sim}$. This is implied by the following expression:

$$
\begin{aligned}
\pi(\theta) &= \int \pi(\theta|y^{sim}) \pi(y^{sim}|\theta^{sim}) \pi(\theta^{sim})dy^{sim} d\theta^{sim} \\
&= \int \pi(\theta|y^{sim}) \pi(y^{sim},\theta^{sim}) dy^{sim} d\theta^{sim}
\end{aligned}
$$

@@

<!-- n error - either in computation, code or overall analysis.  -->

<!-- The simulation if done correctly should produce samples from the prior distribution.  -->

If the posterior samples follows the prior distribution, the rank statistic for a given parameter follows a discrete uniform distribution[^2]. The rank statistic is defined as:

[^2]: Proof of this result in Talts et al (2018).

$$
\begin{aligned}
r = rank(\{\theta_1,\dots , \theta_{L}\}, \theta^{sim}) = \sum_{l=1}^{L}1[\theta_{l} < \theta^{sim}]
\end{aligned}
$$

@@

If the computation is well calibrated and the rank statistics follow a discrete uniform distribution, then the posterior credible intervals have sufficient coverage. That is, one way to describe calibration is: for any percentage interval selected over the posterior samples (for example 90%) then there is a 90\% chance that $\theta^{sim}$ falls within this interval. Another way of saying this is a Bayesian analysis is well calibrated if a 90\% credible interval contains the true parameter in 90\% of the SBC iterations.

## 3.2 Implementation
SBC for Hamiltonian Monte Carlo and the Gaussian approximation are performed with 1000 simulation iterations, 1000 observations in each dataset, and post warmup/burnin samples of 999 (which gives 1000 possible rank statistics). This algorithm is summarised below:

| **1) for sim in 1000 iterations:**
|   2) Draw from prior: $\theta^{sim}\sim\pi (\theta)$
|   3) Simulate dataset with 1000 observations: $y^{sim} \sim p(y|\theta^{sim})$
|   4) Draw 999 posterior samples (post warmup) $\{\theta_1,\dots , \theta_{L}\} \sim p(\theta | y^{sim})$
|   5) Compute rank statistics $r = rank(\{\theta_1,\dots , \theta_{L}\}, \theta^{sim})$

## 3.3 Metrics
The key metrics and diagnostics to compare the performance of these methods are the effective sample size (ESS), rank statistics, and chi-squared test statistics.

@@

A drawback of this approach is there are more parameters than data points in this model. An alternative to visually checking for uniformity of all the parameters is to calculate the chi squared statistics for the counts in each histogram bin. Let $b_j$ be the number of counts and $e_j$ the expected count in bin $j$. Then the chi squared statistic is given by:

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \ifXeTeX
    \usepackage{mathspec} % this also loads fontspec
  \else
    \usepackage{unicode-math} % this also loads fontspec
  \fi
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{lipsum}
\usepackage{setspace}
\onehalfspacing
\linespread{1.5}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Comparison of MCMC algorithms in Stochastic Volatility Models},
  pdfauthor={By Benjamin Wee (32921020); Supervised by Catherine Forbes and Lauren Kennedy},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Comparison of MCMC algorithms in Stochastic Volatility Models}
\author{By Benjamin Wee (32921020) \and Supervised by Catherine Forbes
and Lauren Kennedy}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi

\hypertarget{research-goal}{%
\subsection{1. Research Goal}\label{research-goal}}

@@

\hypertarget{background-and-literature-review}{%
\section{2. Background and Literature
Review}\label{background-and-literature-review}}

\hypertarget{stochastic-volatility}{%
\subsection{2.1 Stochastic Volatility}\label{stochastic-volatility}}

Stochastic volatility models are used in financial econometrics to model
the variance or volatility of returns in a financial instrument. The
return of a financial asset often exhibits patterns in the variance
which can be modeled to improve prediction and estimates of risk or tail
events. Unlike other volatility models such as ARCH or GARCH, stochastic
volatility models treat the variance as a random variable. Furthermore,
these models have a state space representation since the variance is
expressed as a latent stochastic process. State space models can be
difficult to estimate since there is a latent state for every data
observation in the time series, which means there are often more
parameters than data points. These models can be estimated using
frequent frameworks but this research will focus on Bayesian estimation
strategies.

Kim, Shephard and Chib (1998) estimate a discrete time, univariate
stochastic volatility model using Bayesian methods which is outlined
below. \(y_t\) is the mean corrected returns of some asset for equally
spaced intervals t. \(\beta\) is a constant scaling factor which is also
defined as \(exp(\mu / 2)\) representing instantaneous volatility.
\(h_t\) is log volatility, where \(h_1\) is a draw from a stationary
distribution and the state equation \(h_{t+1}\) follows a stationary
process governed by the autoregressive parameter \(\phi\) such that
\(|\phi|<1\). This autoregressive parameter represents the persistence
or ``stickiness'' of log volatility and the dispersion parameter
\(\sigma_{\eta}\) is the constant variance of the states. \(\epsilon_t\)
and \(\eta_t\) are standard normal white noise shocks and are
uncorrelated with each other.

\[
\begin{aligned}
y_t =& \space \beta exp(h_t/2) \epsilon_t \\
h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
h_1 \sim& \space normal\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
\end{aligned}
\]

\[
\begin{aligned}
\epsilon_t \sim& \space normal(0,1) \\
\delta_t \sim& \space normal(0,1)
\end{aligned}
\]

Setting \(\beta=1\), the model can be expressed more succinctly as:

\[
\begin{aligned}
y_t \sim& \space normal(0, exp(h_t/2)) \\ 
h_1 \sim& \space normal \left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
h_{t+1} \sim& \space normal(\mu +\phi(h_t - \mu) , \sigma_{\eta}^2), \space\space t\neq 1\\ 
\end{aligned}
\]

Priors for the static parameters are defined below with conjugate priors
on \(\mu\) and \(\sigma^2\):

\[
\begin{aligned}
\mu \sim& \space normal(0, 10^2) \\
\sigma_{\eta}^2 \sim& \space IG(5/2, (0.01\times 5) / 2) \\
\phi^{\ast} \sim& \space beta(20, 1.5) \\
\phi &=  2\phi^{\ast} - 1
\end{aligned}
\]

The prior on \(\phi\) is a ``stretched'' beta distribution. This is a
beta distribution (as defined on the parameter \(\phi^*\)) which has
been transformed to have support (-1, 1).

\hypertarget{estimation-strategy}{%
\subsubsection{Estimation strategy}\label{estimation-strategy}}

Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic
volatility model using a mix of conjugate posterior distributions,
Metropolis Hastings and the Kalman Filter and smoother\footnote{In this
  research the exact software to apply the simulation smoother is
  unavailable, so a more recent simulation smoother is used which is
  based on the software written by the same author.} (de Jong and
Shephard, 1995).

The standard Kalman Filter and smoother is used to compute the posterior
distribution over the latent states. This requires the state and
measurement equations to be linear and conditionally Gaussian. Since the
relationship between \(y_t\) and \(h_t\) in the measurement equation is
not linear, a transformation is applied by squaring and taking the log
of \(y_t\).

\[
\begin{aligned}
y_t^{*} &= log(y_t^2) \\ 
&= log((\epsilon_t exp(h_t/2))^2) \\
&=  log(exp(h_t)) + log(\epsilon_t^2) \\
&= h_t + log(\epsilon_t^2)  \\
&= h_t + z_t \\
\end{aligned}
\]

Where \(z_t = log(\epsilon_t^2)\) follows a log chi-squared distribution
with mean -1.2704 and variance 4.93. The relationship between \(y_t\)
and \(h_t\) is now linear; however, the error is not Gaussian. Since it
is not simple to sample from this parameterisation of the model, KSC use
a mixture of Gaussians to \textbf{approximate} the first 4 moments of
the log chi squared distribution. This is defined by:

\[
\begin{aligned}
f(z_t) = \sum_{i=1}^{K} q_if_N(z_i|m_i-1.2704, \nu_i^2)
\end{aligned}
\]

Where K is the mixture of normal densities \(f_N\), component
probabilities \(q_i\), mean \(m_i-1.2704\) and variance \(\nu_i^2\).
These parameters were selected using moment matching where they found 7
normal densities with varying mean and variance parameters best
approximated the log chi squared moments. These parameters and weights
can be found in Appendix A.

The model can be sampled via the Kalman Filter and simulation smoother
since the model is now linear and conditionally gaussian. The static
parameters \(\mu\) and \(\sigma^2\) are sampled directly from their
conjugate posterior distributions whereas \(\phi\) is sampled via a
Metropolis Hastings accept/reject procedure. The details can be around
in Appendix B.

\hypertarget{hamiltoninan-monte-carlo}{%
\subsection{2.2 Hamiltoninan Monte
Carlo}\label{hamiltoninan-monte-carlo}}

Since the paper was written, new MCMC algorithms have been developed and
enabled the estimation of richer and more complicated models.
Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has
become widely available for efficiently sampling from sophisticated
models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo,
was developed in the physics literature (Duane et al.~1987) before being
applied in the statistics literature by Radford Neal through his works
in Bayesian Neural Networks (Neal, 1995) and statistical computing (Neal
2011). The algorithm has since become widely available through open
source development projects such as Stan (Stan Development Team, 2023)
and PyMC (2023).

@@

The Stan programming language's implementation of Hamiltonian Monte
Carlo will be used for this study. Stan's default algorithm, the
No-U-Turn Sampler (Hoffman \& Gelman 2014), allows for direct sampling
of the specified stochastic volatility model. Hamiltonian Monte Carlo
allows for sampling of the generative model and can flexibly handle
complicated likelihood functions. This approach will also use the same
priors as specified in the Gaussian mixture approximation.

\hypertarget{research-contribution}{%
\subsection{2.3 Research contribution}\label{research-contribution}}

The goal of this research is to determine which algorithm is better at
estimating the discrete time stochastic volatility model (conditional on
the same priors). My main contribution is to design a simulation study
that compares these algorithms and determines which sampling strategy is
better based on the calibration of posterior estimates and the
efficiency of the MCMC sampler.

\hypertarget{methodology}{%
\section{3. Methodology}\label{methodology}}

\hypertarget{design}{%
\subsection{3.1 Design}\label{design}}

Simulation Based Calibration (SBC) is a technique that checks the
calibration of posterior estimates generated by MCMC algorithms. SBC is
conducted by comparing the distribution of rank statistics to the
discrete uniform distribution which arises when an algorithm is
correctly calibrated. The procedure starts by taking draws from the
prior distribution and creating datasets implied by each draw. Rank
statistics are then calculated on the posterior samples conditional on
the simulated data.

To illustrate this procedure, let \(\theta\) be a parameter and \(y\)
represent the dataset. Start with a single draw from the prior
distribution: \[
\begin{aligned}
\theta^{sim} \sim \pi(\theta)
\end{aligned}
\]

@@

\[
\begin{aligned}
y^{sim} \sim \pi (y|\theta^{sim})
\end{aligned}
\]

@@

\[
\begin{aligned}
\{\theta_1,\dots , \theta_{L}\} \sim \pi (\theta | y^{sim})
\end{aligned}
\]

A key result is that the posterior sample
\(\{\theta_1,\dots , \theta_{L}\}\) will share the same distribution as
the prior samples \(\theta^{sim}\). This is implied by the following
expression:

\[
\begin{aligned}
\pi(\theta) &= \int \pi(\theta|y^{sim}) \pi(y^{sim}|\theta^{sim}) \pi(\theta^{sim})dy^{sim} d\theta^{sim} \\
&= \int \pi(\theta|y^{sim}) \pi(y^{sim},\theta^{sim}) dy^{sim} d\theta^{sim}
\end{aligned}
\]

@@

If the posterior samples follows the prior distribution, the rank
statistic for a given parameter follows a discrete uniform
distribution\footnote{Proof of this result in Talts et al (2018).}. The
rank statistic is defined as:

\[
\begin{aligned}
r = rank(\{\theta_1,\dots , \theta_{L}\}, \theta^{sim}) = \sum_{l=1}^{L}1[\theta_{l} < \theta^{sim}]
\end{aligned}
\]

@@

If the computation is well calibrated and the rank statistics follow a
discrete uniform distribution, then the posterior credible intervals
have sufficient coverage. That is, one way to describe calibration is:
for any percentage interval selected over the posterior samples (for
example 90\%) then there is a 90\% chance that \(\theta^{sim}\) falls
within this interval. Another way of saying this is a Bayesian analysis
is well calibrated if a 90\% credible interval contains the true
parameter in 90\% of the SBC iterations.

\hypertarget{implementation}{%
\subsection{3.2 Implementation}\label{implementation}}

SBC for Hamiltonian Monte Carlo and the Gaussian approximation are
performed with 1000 simulation iterations, 1000 observations in each
dataset, and post warmup/burnin samples of 999 (which gives 1000
possible rank statistics). This algorithm is summarised below:

\textbf{1) for sim in 1000 iterations:}\\
\hspace*{0.333em}\hspace*{0.333em}2) Draw from prior:
\(\theta^{sim}\sim\pi (\theta)\)\\
\hspace*{0.333em}\hspace*{0.333em}3) Simulate dataset with 1000
observations: \(y^{sim} \sim p(y|\theta^{sim})\)\\
\hspace*{0.333em}\hspace*{0.333em}4) Draw 999 posterior samples (post
warmup) \(\{\theta_1,\dots , \theta_{L}\} \sim p(\theta | y^{sim})\)\\
\hspace*{0.333em}\hspace*{0.333em}5) Compute rank statistics
\(r = rank(\{\theta_1,\dots , \theta_{L}\}, \theta^{sim})\)

\hypertarget{metrics}{%
\subsection{3.3 Metrics}\label{metrics}}

The key metrics and diagnostics to compare the performance of these
methods are the effective sample size (ESS), rank statistics, and
chi-squared test statistics.

@@

A drawback of this approach is there are more parameters than data
points in this model. An alternative to visually checking for uniformity
of all the parameters is to calculate the chi squared statistics for the
counts in each histogram bin. Let \(b_j\) be the number of counts and
\(e_j\) the expected count in bin \(j\). Then the chi squared statistic
is given by:
